From Reversible Programs to \\ Univalent Universes and Back

### 2021-07-08 Pi+/paper/discussion.tex
working on sec 8

\paragraph{Computing with Univalence} In HoTT, univalence characterises the path type in the universe as equivalences of
types. The map $\term{idtoeqv} : A \id_{\UU} B \to A \eqv B$ can be easily constructed using path induction. The term
$\term{ua} : A \eqv B \to A \id_{\UU} B$, its computation rule $\term{ua-\beta} : (e : A \eqv B) \to
    \term{idtoeqv}(\term{ua}(e)) \id e$, and its extensionality rule $\term{ua-\eta} : (p : A \id_{\UU} B) \to p ==
    \term{ua}(\term{idtoeqv(p)})$ are generally added as postulates when formalising in Agda. Together, $\term{ua}$ and
$\term{ua}-\beta$ give the full univalence axiom $(A \eqv B) \eqv (A \id_{\UU} B)$.

@@

Our work lies at the intersection of programming language theory, category theory, group theory, rewriting theory, and
formalised mathematics. We review related work in the literature for each topic.

\paragraph{Algebraic Theories} In universal algebra, algebraic theories are used to describe algebraic structures, such
as groups or rings. A specific group or ring is a model of the appropriate algebraic theory. Algebraic theories are
usually \emph{presented} in terms of logical syntax, that is, as first-order theories whose signatures allow only
functional symbols, and whose axioms are universally quantified equations. In his seminal
thesis~\cite{lawvereFUNCTORIALSEMANTICSALGEBRAIC1963}, Lawvere defined a presentation-free categorical notion of
universal algebraic structure, called a Lawvere theory.

\paragraph{Free Symmetric Monoidal Category} The forgetful functor from $\SymMonCat$, the 2-category of (small)
symmetric monoidal categories, strong symmetric monoidal functors, and monoidal natural transformations, to the
2-category $\CatCat$, has a left adjoint giving the free symmetric monoidal category $\FSM[\CCat{C}]$ on a category
$\CCat{C}$. This is a 2-monad on $\CatCat$~\cite{blackwellTwodimensionalMonadTheory1989}, whose algebras are (strict)
symmetric monoidal categories. Its construction is known in the literature~\cite{abramskyAbstractScalarsLoops2005}.
Concretely, the objects of $\FSM[\CCat{C}]$ are given by lists of objects of $\CCat{C}$, that is, a pair $(n:\Nat, A:[n]
    \to \CCat{C}_{0})$. An morphism between $(n,A)$ and $(n,B)$ is given by a pair $(\pi,\lambda)$ where $\pi$ is a
permutation of $[n]$, and $\lambda_{i} : A_{i} \to B_{\pi(i)}$ for $1 \leq i \leq n$. Abstractly, this is given by the
Grothendieck construction $\int F$ of the functor $F : \BFin \to \CatCat$ from the groupoid of finite sets and
bijections to $\CatCat$, assigning each natural number $n$ to the $n$-power $C^{n}$ of $C$, and each permutation on
$[n]$ inducing an endofunctor on $C^{n}$ by action. $\BFin$ is the free symmetric monoidal category (groupoid) on one
generator, $\FSM[\unit]$. The free symmetric monoidal category has been used to study
concurrency~\cite{hylandSymmetricMonoidalSketches2004}, petri nets~\cite{baezCategoriesNets2021}, combinatorial
structures~\cite{fioreCartesianClosedBicategory2008}, quantum mechanics~\cite{abramskyAbstractScalarsLoops2005},
bicategorical models of (differential) linear logic~\cite{melliesTemplateGamesDifferential2019}.

\paragraph{Curry-Howard-Lambek correspondence} In~\citet{curryCurryEssaysCombinatory1980}, Lambek extended the
Curry-Howard correspondence to cartesian-closed categories. In this work, we have established a correspondence between a
fragment of the $\PiLang$ family of reversible programming languages and symmetric monoidal groupoids. The Curry-Howard
part of this correspondence with reversible logic was established in~\cite{sparksSuperstructuralReversibleLogic2014}.

\paragraph{Rewriting} We presented a rewriting system for the Coxeter relations for $\Sn$ to solve its word problem.

\paragraph{Computational group theory} Coxeter relations are used in computational group theory to study XXX problems.

\paragraph{Univalent Fibrations} Univalent Fibrations were introduced by~\citet*{kapulkinUnivalenceSimplicialSets2018},
to build a model of Voevodsky's \emph{univalence} principle in simplicial sets.
\citet{christensenCharacterizationUnivalentFibrations2015} studied characterisations of univalent fibrations using the
$\BAut$ construction. Univalent typoids~\cite{petrakisUnivalentTypoids2019a} are a different presentation of univalent
subuniverses.

@@

\paragraph{Extensional vs. Intensional View of Permutations.}
Although the meaning of a permutation - that is, the specific bijective function
that it represents - is in some sense all there is, manipulating them
syntactically still has its advantages. By writing reversible programs, we think
of them in an intensional way. Comparing two programs for equality by evaluating
them on all points in the domain is a very crude - and maybe even inefficient -
way of doing that. ~\cite{Kuehlmann:2006:RBR:2298470.2300327,10.1007/978-3-540-24605-3_4,Yamashita:2010:FEQ:1835957.1835965}.

>>>>

The main theme of our work is the semantic foundation of reversible languages. We prove that a programming language
presentation of reversible programming based on algebraic types matches---exactly---the categorified group-theoretic
semantics, thereby closing the circle on a complete Curry-Howard-Lambek correspondence for reversible
languages. Historically, the first such correspondence was between the $\lambda$-calculus, intuitionistic logic, and
cartesian-closed categories~\cite{curryCurryEssaysCombinatory1980}. For reversible languages, the Curry-Howard
correspondence was established by~\citet{sparksSuperstructuralReversibleLogic2014} and the Lambek correspondence is
established in this work.

In the remainder of this section, we discuss some broader related work.

%%%%%%%%%%%%%%
\paragraph{Rewriting} We presented a rewriting system for the Coxeter relations for $\Sn$ to solve its word problem.

%%%%%%%%%%%%%%
\paragraph{Extensional vs. Intensional View of Permutations.}
Although the meaning of a permutation - that is, the specific bijective function
that it represents - is in some sense all there is, manipulating them
syntactically still has its advantages. By writing reversible programs, we think
of them in an intensional way. Comparing two programs for equality by evaluating
them on all points in the domain is a very crude - and maybe even inefficient -
way of doing that. ~\cite{Kuehlmann:2006:RBR:2298470.2300327,10.1007/978-3-540-24605-3_4,Yamashita:2010:FEQ:1835957.1835965}.

%%%%%%%%%%%%%%
\paragraph{Free Symmetric Monoidal Category} The forgetful functor from $\SymMonCat$, the 2-category of (small)
symmetric monoidal categories, strong symmetric monoidal functors, and monoidal natural transformations, to the
2-category $\CatCat$, has a left adjoint giving the free symmetric monoidal category $\FSM[\CCat{C}]$ on a category
$\CCat{C}$. This is a 2-monad on $\CatCat$~\cite{blackwellTwodimensionalMonadTheory1989}, whose algebras are (strict)
symmetric monoidal categories. Its construction is known in the literature~\cite{abramskyAbstractScalarsLoops2005}.
Concretely, the objects of $\FSM[\CCat{C}]$ are given by lists of objects of $\CCat{C}$, that is, a pair $(n:\Nat, A:[n]
    \to \CCat{C}_{0})$. An morphism between $(n,A)$ and $(n,B)$ is given by a pair $(\pi,\lambda)$ where $\pi$ is a
permutation of $[n]$, and $\lambda_{i} : A_{i} \to B_{\pi(i)}$ for $1 \leq i \leq n$. Abstractly, this is given by the
Grothendieck construction $\int F$ of the functor $F : \BFin \to \CatCat$ from the groupoid of finite sets and
bijections to $\CatCat$, assigning each natural number $n$ to the $n$-power $C^{n}$ of $C$, and each permutation on
$[n]$ inducing an endofunctor on $C^{n}$ by action. $\BFin$ is the free symmetric monoidal category (groupoid) on one
generator, $\FSM[\unit]$. The free symmetric monoidal category has been used to study
concurrency~\cite{hylandSymmetricMonoidalSketches2004}, petri nets~\cite{baezCategoriesNets2021}, combinatorial
structures~\cite{fioreCartesianClosedBicategory2008}, quantum mechanics~\cite{abramskyAbstractScalarsLoops2005},
bicategorical models of (differential) linear logic~\cite{melliesTemplateGamesDifferential2019}.

%%%%%%%%%%%%%%
\paragraph{Computing with Univalence} In HoTT, univalence characterises the path type in the universe as equivalences of
types. The map $\term{idtoeqv} : A \id_{\UU} B \to A \eqv B$ can be easily constructed using path induction. The term
$\term{ua} : A \eqv B \to A \id_{\UU} B$, its computation rule $\term{ua-\beta} : (e : A \eqv B) \to
    \term{idtoeqv}(\term{ua}(e)) \id e$, and its extensionality rule $\term{ua-\eta} : (p : A \id_{\UU} B) \to p ==
    \term{ua}(\term{idtoeqv(p)})$ are generally added as postulates when formalising in Agda. Together, $\term{ua}$ and
$\term{ua}-\beta$ give the full univalence axiom $(A \eqv B) \eqv (A \id_{\UU} B)$.

@@

%%%%%%%%%%%%%%
\paragraph{Univalent Fibrations} Univalent Fibrations were introduced by~\citet*{kapulkinUnivalenceSimplicialSets2018},
to build a model of Voevodsky's \emph{univalence} principle in simplicial sets.
\citet{christensenCharacterizationUnivalentFibrations2015} studied characterisations of univalent fibrations using the
$\BAut$ construction. Univalent typoids~\cite{petrakisUnivalentTypoids2019a} are a different presentation of univalent
subuniverses.

@@

%%%%%%%%%%%%%%
\paragraph{Algebraic Theories} In universal algebra, algebraic theories are used to describe algebraic structures, such
as groups or rings. A specific group or ring is a model of the appropriate algebraic theory. Algebraic theories are
usually \emph{presented} in terms of logical syntax, that is, as first-order theories whose signatures allow only
functional symbols, and whose axioms are universally quantified equations. In his seminal
thesis~\cite{lawvereFUNCTORIALSEMANTICSALGEBRAIC1963}, Lawvere defined a presentation-free categorical notion of
universal algebraic structure, called a Lawvere theory.

% %%%%%%%%%%%%%%
% \paragraph{Computational group theory} Coxeter relations are used in computational group theory to study XXX problems.

% \todo{but computation is usually done by using Coxeter matrices~\cite{davisGeometryTopologyCoxeter2008}.}
